QuickSched: Task-based parallelism with dependencies and conflicts

### 2013-10-29 paper/paper.tex
add new figures, backup of text.

% My own macros
\usepackage{epsfig}
\usepackage{listings}
\lstset{%
    language=C,
    basicstyle=\small\tt,
    numbers=left,
    numberstyle=\tiny
    }
\newcommand{\oh}[1]
    {\mbox{$ {\mathcal O}( #1 ) $}}
\newcommand{\eqn}[1]
    {(\ref{eqn:#1})}
\newcommand{\fig}[1]
    {Figure~\ref{fig:#1}}
\newcommand{\tabl}[1]
    {Table~\ref{tab:#1}}
\newcommand{\figs}[2]
    {Figures~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\sect}[1]
    {Section~\ref{sec:#1}}
\newcommand{\cuda}
    {{\sc cuda}\xspace}
\newcommand{\namd}
    {{\sc namd}\xspace}
\newcommand{\acemd}
    {{\sc acemd}\xspace}
\newcommand{\fenzi}
    {{\sc fenzi}\xspace}
\newcommand{\mdcore}
    {{\tt mdcore}\xspace}
\newcommand{\swift}
    {{\sc swift}\xspace}
\newcommand{\bsf}[1]
    {\textbf{\textsf{#1}}}

@@

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task {
    int type, wait;
    void *data;
    struct task **unlocks;
    struct resource **locks, **uses;
    int size_data, nr_unlocks, nr_locks, nr_uses;
    int cost, weight;
    };
    \end{lstlisting}
\end{minipage}\end{center}

@@

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run ( qsched *s , void (*fun)( int , void * ) ) {
    qsched_start( s );
    #pragma omp parallel
    {
        int qid = omp_get_thread_num() % s->nr_queues;
        struct task *t;
        while ( ( t = qsched_gettask( s , qid ) ) != NULL ) {
            fun( t->type , t->data );
            qsched_done( s , t );
            }
        }
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \cite{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~12 is executed
concurrently.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via the function {\tt qsched\_gettask}
and its type and data are passed to a user-supplied execution
function.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qstack\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.

At the start of a parallel computation, {\tt qsched\_start}
identifies the tasks that have no dependencies and sends them
to queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_start ( qsched *s ) {
    set the task waits and weights.
    s->waiting = s->nr_tasks;
    for ( k = 0 ; k < s->nr_tasks && s->tasks[tid[k]].wait == 0 ; k++ )
        qsched_enqueue( s , &s->tasks[tid[k]] );
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the operations in line~2 are described in \sect{tasks}.

@@

The function {\tt qsched\_gettask} fetches a task from
one of the queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_gettask ( qsched *s , int qid ) {
    struct task *res = NULL;
    int k;
    while ( s->waiting ) {
        if ( ( res = queue_get( s->queues[qid] ) ) == NULL ) {
            k = id of queue with largest count.
            if ( ( res = queue_get( s->queues[k] ) ) != NULL )
                break;
            }
        else
            break;
        }
    if ( res != NULL ) {
        for ( k = 0 ; k < res->nr_locks ; k++ )
            res->locks[k]->owner = qid;
        for ( k = 0 ; k < res->nr_uses ; k++ )
            res->uses[k]->owner = qid;
        }
    return res;
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the parameter {\tt qid} is the index of the
prefered queue.
If the queue is empty, or all of the tasks in that queue had
unresolved conflicts, the scheduler uses {\em work stealing}
\cite{ref:Blumofe1999}, i.e.~it looks for the queue with the
largest task count, and attempts to steal a task from it
(line~7).
If a task could be obtained from any queue (line~13),
the resources it locks and uses are marked as now being owned
by the prefered queue (lines~14--17).
Finally, the task, or {\tt NULL} if no task could be obtained,
is returned.

@@

@@

@@

@@

@@

Main points:
\begin{itemize}
    \item Conflicts are better than dependencies.
    \item Specifying the entire task tree before execution
        gives better scheduling.
    \item Combination of locality, weight, and availability.
\end{itemize}

@@

% Bibliography
\bibliographystyle{hacm}
\bibliography{quicksched}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

>>>>

% My own macros
\usepackage{epsfig}
%\usepackage{hyperref}
\usepackage{listings}
\lstset{%
    language=C,
    basicstyle=\small\tt,
    numbers=left,
    numberstyle=\tiny
    }
\newcommand{\oh}[1]
    {\mbox{$ {\mathcal O}( #1 ) $}}
\newcommand{\eqn}[1]
    {(\ref{eqn:#1})}
\newcommand{\fig}[1]
    {Figure~\ref{fig:#1}}
\newcommand{\tabl}[1]
    {Table~\ref{tab:#1}}
\newcommand{\figs}[2]
    {Figures~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\sect}[1]
    {Section~\ref{sec:#1}}
\newcommand{\cuda}
    {{\sc cuda}\xspace}
\newcommand{\namd}
    {{\sc namd}\xspace}
\newcommand{\acemd}
    {{\sc acemd}\xspace}
\newcommand{\fenzi}
    {{\sc fenzi}\xspace}
\newcommand{\mdcore}
    {{\tt mdcore}\xspace}
\newcommand{\swift}
    {{\sc swift}\xspace}
\newcommand{\bsf}[1]
    {\textbf{\textsf{#1}}}

@@

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task {
  int type, wait;
  void *data;
  struct task **unlocks;
  struct resource **locks, **uses;
  int size_data, nr_unlocks, nr_locks, nr_uses;
  int cost, weight;
  };
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the {\tt data}, {\tt unlocks}, {\tt locks},
and {\tt uses} arrays are pointers to the contents of other
arrays, i.e.~they are not allocated individually.

@@

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run ( qsched *s , void (*fun)( int , void * ) ) {
  qsched_start( s );
  #pragma omp parallel
  {
    int qid = omp_get_thread_num() % s->nr_queues;
    struct task *t;
    while ( ( t = qsched_gettask( s , qid ) ) != NULL ) {
      fun( t->type , t->data );
      qsched_done( s , t );
      }
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \cite{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~12 is executed
concurrently.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via the {\em execution function}
{\tt qsched\_gettask}
and its type and data are passed to a user-supplied execution
function.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qstack\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.

At the start of a parallel computation, {\tt qsched\_start}
identifies the tasks that have no dependencies and sends them
to queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_start ( qsched *s ) {
  sort out pointers in tasks.
  set the task waits and weights.
  s->waiting = s->nr_tasks;
  for ( k = 0 ; k < s->nr_tasks && s->tasks[tid[k]].wait == 0 ; k++ )
    qsched_enqueue( s , &s->tasks[tid[k]] );
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where line~2 sets the {\tt unlocks}, {\tt locks},
and {\tt uses} pointers in the tasks.
The operations in line~3 are described in \sect{tasks}.

@@

The function {\tt qsched\_gettask} fetches a task from
one of the queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_gettask ( qsched *s , int qid ) {
  struct task *res = NULL;
  int k;
  while ( s->waiting ) {
    if ( ( res = queue_get( s->queues[qid] ) ) == NULL ) {
      loop over all other queues in random order with index k
        if ( ( res = queue_get( s->queues[k] ) ) != NULL )
          break;
      }
    else
      break;
    }
  if ( res != NULL ) {
    for ( k = 0 ; k < res->nr_locks ; k++ )
      res->locks[k]->owner = qid;
    for ( k = 0 ; k < res->nr_uses ; k++ )
      res->uses[k]->owner = qid;
    }
  return res;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the parameter {\tt qid} is the index of the
prefered queue.
If the queue is empty, or all of the tasks in that queue had
unresolved conflicts, the scheduler uses {\em work stealing}
\cite{ref:Blumofe1999}, i.e.~it loops over all other queues
in a random order (line~6) and tries to get a task from them
(line~7).
If a task could be obtained from any queue (line~13),
the resources it locks and uses are marked as now being owned
by the prefered queue (lines~14--17).
Finally, the task, or {\tt NULL} if no task could be obtained,
is returned.

@@

The algorithms, as described in the previous section, have
all been implemented as part of the Open-Source C-language
QuickSched library.\footnote{\url{http://sourceforge.net/projects/quicksched/}}
This section describes the interface functions and how they
are called, whereas the following section contains
examples of how QuickSched can be used.

As mentioned previously, the {\tt qsched} object is the main
interface to the task scheduler.
As such, it provieds functionality for task and resource
creation, for assigning resources to tasks, either as locks
or uses, and for assigning dependencies between tasks.
The tasks and resources themselves are opaque to the
user, and handles of the types {\tt qsched\_task\_t}
and {\tt qsched\_res\_t} are used instead.

The main functions for setting up the scheduler are:
\begin{itemize}
    \item {\tt void qsched\_init( struct qsched *s , int nr\_queues , int flags )} \\
        Initializes a {\tt qsched} object with the given number of queues.
        The {\tt flags} parameter can be set to any bitwise or combination
        of {\tt qsched\_flag\_none},
        {\tt qsched\_flag\_yield}, and {\tt qsched\_flag\_pthread},
        which are described further below.
        This function must be called before any of the other
        functions are used.
        \vspace{1mm}
    \item {\tt void qsched\_free( struct qsched *s )} \\
        Releases all the memory and other resources allocated by the
        given {\tt qsched} object.
        After this function has been called, the {\tt qsched} will
        need to be re-initialized for reuse.
        \vspace{1mm}
    \item {\tt void qsched\_reset( struct qsched *s )} \\
        Clears the tasks and resources in the given {\tt qsched},
        but does not release the allocated memory or change
        the number of queues.
        \vspace{1mm}
    \item {\tt qsched\_task\_t qsched\_addtask( struct qsched *s , int type , unsigned int flags , void *data , int data\_size , int cost )} \\
        Creates a new task within the given {\tt qsched} and returns
        its handle.
        The {\tt type} and {\tt data} field are copied into the {\tt qsched}
        and passed to the execution function when the {\tt qsched} is run.
        The parameter flags is either {\tt task\_flag\_none} or
        {\tt task\_flag\_virtual}.
        Tasks marked as virtual do not have any action associated with them,
        e.g. they are used only to group or otherwise dependencies, and
        are not passed to the execution function in {\tt qsched\_run}.
        \vspace{1mm}
    \item {\tt qsched\_res\_t qsched\_addres( struct qsched *s , qsched\_res\_t parent )} \\
        Creates a new resource within the given {\tt qsched} and returns
        its handle.
        The {\tt parent} field is the handle of the heirarchical parent of
        the new resource or {\tt qsched\_res\_none} if the resource
        has no hierarchical parent.
        \vspace{1mm}
    \item {\tt void qsched\_addlock( struct qsched *s , qsched\_task\_t t , qsched\_res\_t res )} \\
        Append the resource {\tt res} to the task {\tt t}'s list of
        locks.
        The task {\tt t} will then conflict with any other task that
        also locks the resource {\tt res}, its hierarchical parents, or
        any resource hierarchically below it.
        \vspace{1mm}
    \item {\tt void qsched\_adduse( struct qsched *s , qsched\_task\_t t , qsched\_res\_t res )} \\
        Similar to {\tt qsched\_addlock}, yet the resource is only used and
        is not part of a conflict.
        This information is used when assinging tasks to specific queues.
        \vspace{1mm}
    \item {\tt void qsched\_addunlock( struct qsched *s , qsched\_task\_t ta , qsched\_task\_t tb )} \\
        Appends the task {\tt tb} to the list of tasks that the task {\tt ta}
        unlocks.
        The task {\tt tb} then depends on the task {\tt ta}.
        \vspace{1mm}
    \item {\tt void qsched\_run( struct qsched *s , int nr\_threads , qsched\_funtype fun )} \\
        Executes the tasks in the given {\tt qsched} using {\tt nr\_threads}
        threads via the execution function {\tt fun}, as described in the
        previous section.
        Once a {\tt qsched} has been initialized and filled with
        tasks and resources, it can be run more than once.
        \vspace{1mm}
\end{itemize}

The library can be compiled to use either OpenMP or the
{\tt pthreads} library \cite{ref:Pthreads1995}.
OpenMP is the default, but calling {\tt qsched\_init} with
either the {\tt qsched\_flag\_yield}
or the {\tt qsched\_flag\_pthread} switches to using {\tt pthreads}
for the parallel loop.

OpenMP has the advantage of being available for most compilers
and also potentially providing some extra platform-specific
scheduling features, e.g.~optimal thread location and/or affinity,
and of integrating seamlessly with other parallel parts of the
user application.
The disadvantage of using OpenMP is that it does not provide
any mechanism for yielding a thread if no tasks are available,
i.e. the main loop in {\tt qsched\_gettask}, described in the
previous section, will spin until a task becomes available.
This may be a problem if other parts of the user application
are running in the background simultaneously.
Calling {\tt qsched\_init} with the {\tt qsched\_flag\_yield}
forces the use of {\tt pthreads} and uses conditional variables
to wait for a new task to be enqueued if obtaining a task
from any of the queues fails.
This relinquishes the waiting computational thread for other
processes.

@@

This section presents two test cases showing both
how QuickSched can be used in real-world applications, and
providing benchmarks to assess its efficiency and scalability.

The source code of both examples is distributed with the
QuickSched library.

@@

\citeN{ref:Buttari2009} introduced the concept of using task-based
parallelsim for tile-based algorithms in numerical linear algebra,
presenting parallel codes for the Cholesky, LU, and QR
decompositions.
These algorithms are now part of the PLASMA and MAGMA
libraries for parallel linear algebra \cite{ref:Agullo2009}.
The former uses the QUARK task scheduler, which was originally
designed for it, while the latter currently uses the StarPU
task scheduler.

\begin{figure}
    \centerline{\epsfig{file=figures/QR.pdf,width=0.8\textwidth}}
    \caption{Task-based QR decomposition of a matrix consisting
        of $4\times 4$ tiles.
        Each circle represents a tile, and its colour represents
        the type of taks on that tile at that level.
        Empty circles have no task associated with them.
        The arrows represent dependencies at each level, and
        tasks at each level also implicitly depend on the
        task at the same location in the previous level.}
    \label{fig:QR}
\end{figure}

The tiled QR factorization is based on four basic tasks,
or kernels, as shown in \fig{QR}.
For a matrix consisting of $N\times N$ tiles, $N$ passes,
or levels, are computed, each computing a column and row of the QR
decomposition.
The tasks can be defined in terms of the tuples $(i,j,k)$,
where $i$ and $j$ are the row and column of the tile, respectively,
and $k$ is its level:

\begin{center}
    \begin{tabular}{llll}
        Task & cond. & depends on task & locks tile \\
        \hline
        \epsfig{file=figures/TaskRed.pdf,height=9pt} DGEQRF & $i=j=k$ & $(i,j,k-1)$ & $(i,j)$ \\
        \epsfig{file=figures/TaskGreen.pdf,height=9pt} DLARFT & $i=k$, $j>k$ & $(i,j,k-1)$, $(k,k,k)$ & $(i,j)$ \\
        \epsfig{file=figures/TaskBlue.pdf,height=9pt} DTSQRF & $i>k$, $j=k$ & $(i,j,k-1)$, $(i-1,j,k)$ & $(i,j)$, $(j,j)$ \\
        \epsfig{file=figures/TaskYellow.pdf,height=9pt} DSSRFT & $i>k$, $j>k$ & $(i,j,k-1)$, $(k,j,k)$, $(i,k,k)$ & $(i,j)$ \\
        \hline
    \end{tabular}
\end{center}

\noindent where the task names are the BLAS-like operation
performed on the given tiles.
Every task depends on the task at the same position and the
previous level, i.e.~the task $(i,j,k)$ always depends on
$(i,j,k-1)$ for $k>1$.
Each task also modifies its own tile $(i,j)$, and the DTSQRF
task additionally modifies the lower triangular part of the $(j,j)$th tile.

Setting up the dependencies and locks for a matrix of
$m\times n$ tiles is implemented as shown in \fig{CodeQR},

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
enum { tDGEQRF , tDLARFT , tDTSQRF , tDSSRFT };
void make_tasks ( struct qsched *s , int m , int n ) {
  int i, j, k, data[3];
  qsched_task_t tid[ m*n ], tid_new;
  qsched_res_t rid[ m*n ];
  for ( k = 0 ; k < m*n ; k++ ) {
    tid[k] = qsched_task_none;
    rid[k] = qsched_addres( s , qsched_res_none );
    }
  for ( k = 0 ; k < m && k < n ; k++ ) {
    /* DGEQRF task at (k,k). */
    data[0] = k; data[1] = k; data[2] = k;
    tid_new = qsched_addtask( s , tDGEQRF , qsched_flags_none ,
      data , sizeof(int)*3 , 2 );
    qsched_addlock( s , tid_new , rid[ k*m + k ] )
    if ( tid[ k*m + k ] != qsched_task_none )
      qsched_addunlock( s , tid[ k*m + k ] , tid_new );
    tid[ k*m + k ] = tid_new;
    for ( j = k+1 ; j < n ; j++ ) {
      /* DLARFT task at (k,j). */
      data[0] = k; data[1] = j; data[2] = k;
      tid_new = qsched_addtask( s , tDLARFT , qsched_flags_none ,
        data , sizeof(int)*3 , 3 );
      qsched_addlock( s , tid_new , rid[ j*m + k ] )
      qsched_addunlock( s , tid[ k*m + k ] , tid_new );
      if ( tid[ j*m + k ] != qsched_task_none )
        qsched_addunlock( s , tid[ j*m + k ] , tid_new );
      tid[ j*m + k ] = tid_new;
      }
    for ( i = k+1 ; i < m ; i++ ) {
      /* DTSQRF task at (i,k). */
      data[0] = i; data[1] = k; data[2] = k;
      tid_new = qsched_addtask( s , tDTSQRF , qsched_flags_none ,
        data , sizeof(int)*3 , 3 );
      qsched_addlock( s , tid_new , rid[ k*m + i ] )
      qsched_addlock( s , tid_new , rid[ k*m + k ] )
      qsched_addunlock( s , tid[ k*m + k ] , tid_new );
      if ( tid[ k*m + i ] != qsched_task_none )
        qsched_addunlock( s , tid[ k*m + i ] , tid_new );
      tid[ k*m + i ] = tid_new;
      for ( j = k+1 ; j < n ; j++ ) {
        /* DSSRFT task at (i,j). */
        data[0] = i; data[1] = j; data[2] = k;
        tid_new = qsched_addtask( s , tDSSRFT , qsched_flags_none ,
          data , sizeof(int)*3 , 5 );
        qsched_addlock( s , tid_new , rid[ j*m + i ] )
        qsched_addunlock( s , tid[ j*m + k ] , tid_new );
        qsched_addunlock( s , tid[ k*m + i ] , tid_new );
        if ( tid[ j*m + i ] != qsched_task_none )
          qsched_addunlock( s , tid[ j*m + i ] , tid_new );
        tid[ j*m + i ] = tid_new;
        }
      }
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
\caption{Example code to generate the tasks for the tiled
    QR decomposition.}
\label{fig:CodeQR}
\end{figure}

Setting up the dependencies and locks for a matrix of
$m\times n$ tiles is implemented as shown in \fig{CodeQR},
where the $m\times n$ matrix {\tt tid} stores the handles
of the last task at position $(i,j)$ and is initialized with
empty tasks (line~7).
Similarly, {\tt rid} stores the handles of the resources for each
tile of the matrix, allocated in line~8.

The following loops mirror the task generation described in
Algorithm~2 of \citeN{ref:Buttari2009}.
E.g.~for each level {\tt k} (line~10), a DGEQRF task is created
for tile $(k,k)$ (lines~13--14).
A lock is added for the newly created task on the
resource associated with the $(k,k)$th tile (line~15).
If a task exists at that position at the previous level
(line~16), a dependency is added between the old task and
the new (line~17), and the new task is stored in {\tt tid}
(line~18).
The remaining tasks are generated in the same way, with
their respective locks and dependencies.

The execution function for these tasks simply calls the apropriate
kernels on the matrix tiles given by the task data:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun ( int type , void *data ) {
  int *idata = (int *)data;
  int i = idata[0], j = idata[1], k = idata[2];
  switch ( type ) {
    case tDGEQRF:
      DGEQRF( A[i,j] , ... );
      break;
    case tDLARFT:
      DLARFT( A[i,j] , A[i,i] , ... );
      break;
    case tDTSQRF:
      DTSQRF( A[i,j] , A[j,j]... );
      break;
    case tDSSRFT:
      DSSRFT( A[i,j] , A[i,k] , A[k,j] , ... );
      break;
    default:
      error( "Unknown task type." );
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt A} is the matrix over which the QR
decmposition is executed.

@@

The Barnes-Hut tree-code is an algorithm to compute the
interactions between a set of particles

@@

Main points/differences to other schedulers:
\begin{itemize}
    \item Conflicts.
    \item Specifying the entire task tree before execution
        gives better scheduling.
    \item Combination of locality, weight, and availability.
    \item Compact implementation.
\end{itemize}

Possible improvements
\begin{itemize}
    \item Use number of locks as well when selecting from a queue.
    \item Use total/max queue weight when selecting for work-stealing.
    \item Use different resource sizes when computing queue score.
    \item Dynamically add new tasks, but potentially wrecks havoc
        with the weights.
\end{itemize}

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{quicksched}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012
